---
title: Viewing source code for a function
author: Pawel Sobala
date: '2020-05-29'
slug: viewing-source-code-for-a-function
categories:
  - R
  - programming
  - fucntion
tags:
  - R
description: Desc
hacker_news_id: ''
lobsters_id: ''
meta_img: /images/image.jpg
---

# Rationale

Yesterday, I wanted to take a look at the source code of `cor.test()`{.r} function in R. I did what I usually do in such case - namely, I called the function and I expected to see the code straight way. 

```{r example_cor_test}
cor.test
```

Unfortunately, I didn't get the code but instead I recieved  information that it comes from a S3 generic function. More on this topic can be found in a great book "Advanced R" by Hadley Wickham, which is available for free on his [website](https://adv-r.hadley.nz/s3.html). I forgot, how to perform it, so it is always good to review the basics. 

A great post on the overall topic of viewing source code is available on the [StackOverflow](https://stackoverflow.com/questions/19226816/how-can-i-view-the-source-code-for-a-function). This thread includes topics like:
* The S3 method dispatch system
* The S4 method dispatch system
* Functions that call unexported functions
* Functions that call compiled code
* Compiled code in a package
* Compiled code in a base package
* Compiled code built into the R interpreter
* Infix Operators

At the end of this post I will highlight the most useful ideas in my opinion that are mentioned in this thread.

The purpose of this post is to document, how one can quickly and efficiently investigate the underlying code of the `cor.test()`{.r} function.

# Viewing source code

As already noted, the `cor.test`{.r} function is a S3 generic function. For S3 classes, one can use the `methods`{.r} function to investigate it.

```{r}
methods("cor.test")
```

It turns out, one needs to call ``{.r} instead of `cor.test`{.r}. Please note that in the following R line `:::`{.r} is used to view the source code.

```{r}
head(stats:::cor.test.default)
```

Perfect, it works. Please note, that I used additionally the `head`{.r} function to print only the first 6 lines of the script. 

## Useful tricks

As mentioned in the introduction, below are listed functions that in my opinion are the most commonly used tricks that are mentioned in this [post](https://stackoverflow.com/questions/19226816/how-can-i-view-the-source-code-for-a-function) on StackOverflow.

```{r}
# Opening in RStudio a new window with a function:
# invisible(edit(FUNCTION))
# 
# Function to find the source of a given function:
# getAnywhere(FUNCTION)
# 
# Opening function in a new r script (outside RStudio):
# edit(getAnywhere('rfcv'), file='source_rfcv.r')
# 
# Opening function in a new script in RStudio:
# Just click on the function keeping F2 pressed
```


