---
title: "How to find packages on CRAN from RStudio"
description: Using librarian package to find packages on CRAN directly from RStudio
author: "Pawel Sobala"
date: '2020-02-10'
slug: how-to-find-packages-on-cran-from-rstudio
categories: ["R", "Packages"]
tags: ["R Packages", "CRAN", "Librarian"]
#output: 
#  html_document:
#    keep_md: true
#    toc: true
#    toc_depth: 3
#    toc_float: 
#      collapsed: false
#      smooth_scroll: true
#    number_sections: true  ## if you want number sections at each table header
#    theme: united  # many options for theme, this one is my favorite.
#    highlight: tango  # specifies the syntax highlighting style - tango # https://www.garrickadenbuie.com/blog/pandoc-syntax-highlighting-examples/
---
<style>
body {
text-align: justify}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
library(librarian)
library(tidyverse)
```

# Motivation

When writing a new blog post on the assessment of quality of fitted Self-Organizing Map (SOM) to data I stumbled upon a problem on how to find all R packages that are related to a particular subject on CRAN. In particular, my problem was to find all packages that might contain any items related to the assessment of SOM.

# Librarian Package

After googling for a while I found the `library(librarian)`{.r} package which contains a function `browse_cran()`{.r} to search package names and descriptions within CRAN. For example, if you can’t remember what the “colorbrewer” package is actually called, you just type `browse_cran("colorbrewer")`{.r}.

I decided to test it and to find any packages that are related to the Self Organizing Maps topic.
I found out that I cannot use the argument `som` to find any items as all words containing `some` are also returned. A quote `organizing map` should work.

```{r librarian_browse_cran}
som_data <- browse_cran("organizing map")
```


Indeed, it returns a couple of packages that might have some functions to work with Self-Organizing Maps!

# How does the function work?

Let's check the contents of the function.

```{r browse_cran_function_contents}
browse_cran
```

Actually, there is a function `CRAN_package_db()`{.r} from the `library(tools)`{.r} which obtains information about current packages in the CRAN package repository, and their check status.

```{r}
tools::CRAN_package_db
```

Therefore, it seems that all package info is stored on CRAN in one repository. Let's check this hypothesis, if any time stamp is available in the function ``{.r}.

```{r}
db <-  tools::CRAN_package_db()
colnames(db)
```

Actually there are `64` different variables in this data object from CRAN. 
The last 20 available dates are following:

```{r, warning=FALSE}
date_db <- db$Date %>% as.Date()
date_db %>% 
  sort(na.rm = TRUE) %>% 
  tail(20)
```

There are some outliers but as the lastest sensible dates are from the same day when this analysis is created then it is safe to assume that `packages.rds`{.r} file is up-to-date. Hence, this approach can be used to find out more about R Packages in CRAN directly from RStudio!