---
title: "Calculating YTD return of EM curencies"
description: "Analysis of returns of Emerging Markets currencies in 2020"
author: "Pawel Sobala"
date: "2020-05-03"
slug: analysis-of-em-curencies
categories: ["R", "programming", "analysis", "EM"]
tags: ["R", "analysis"]
output: html_document
#output: 
#  html_document:
#    keep_md: true
#    toc: true
#    toc_depth: 3
#    toc_float: 
#      collapsed: false
#      smooth_scroll: true
#    number_sections: true  ## if you want number sections at each table header
#    theme: united  # many options for theme, this one is my favorite.
#    highlight: tango  # specifies the syntax highlighting style - tango # https://www.garrickadenbuie.com/blog/pandoc-syntax-highlighting-examples/

---

<style>
body {
text-align: justify}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

setwd("C:/Users/pawel/Documents/R/New_posts")
rm(list=ls())

library(data.table)
library(purrr)
library(tidyverse)
library(reshape2)
library(lubridate)
require(quantmod)

options(scipen=999)

read_data_stooq <- function(tickers, 
                            join_type = full_join, 
                            apply_approx = TRUE) {
  data_all <- purrr::map(tickers, read_rawdata_stooq)
  
  merged_data <- merge_stooq_data(data_all, join_type)
  if (apply_approx == TRUE) {
    merged_data <-apply_approx(merged_data)
  }
  merged_data
}

read_rawdata_stooq <- function(x) {
  fread(paste0("https://stooq.pl/q/d/l/?s=", x, "&i=d"), fill = TRUE) %>%
    select(Date = Data, !!x := Zamkniecie)
} #' zamkniecie' means close value

multi_join <- function(list_of_loaded_data, join_func, ...) {
  output <- Reduce(function(x, y) {
    join_func(x, y, ...)
  }, list_of_loaded_data)
}

approx_ts <- function(x) {
  na.approx(x, maxgap = 10, rule = 2:2)
}

merge_stooq_data <- function(input_data = data_all,
                             join_type = full_join) {
  merged_data <- multi_join(input_data, join_type) %>%
    mutate(Date = as.Date(Date)) %>%
    arrange(Date)
}

apply_approx <- function(input_data){
  input_data_approx <- purrr::map_df(
    input_data[, -1],
    approx_ts
  )
  input_data <- cbind(
    Date = input_data[, 1],
    input_data_approx
  )
}

generate_returns <- function(input_data, na.rm=FALSE,return_type = "log") {
  temp_colnames <- input_data %>% colnames()
  if(na.rm==TRUE){
    input_data <- na.omit(input_data)
  }
  
  if (return_type == "log") {
    input_data <- data.frame(
      cbind.data.frame(
        Date = input_data[-1, 1],
        diff(as.matrix(log(input_data[, -1])))
      )
    )
  } else {
    input_data <- data.frame(
      cbind.data.frame(
        Date = input_data[-1, 1],
        diff(as.matrix(input_data[, -1])) / input_data[-nrow(input_data), -1]
      )
    )
  }
  colnames(input_data) <- temp_colnames
  input_data
}

```

# Introduction

The goal of this short article is to calcuate the returns of Emerging Markets (EM) currencies in 2020.
The main observation is that all of those currencies since mid-Febraury 2020 (Coronavirus 2020 Financial Crisis) significantly weakened in comparison to EUR or USD.
The question is also, was there any curency that performed better in EUR than in USD?

## Specification of Currencies

The currencies that are going to be analyzed as a part of this analysis has been chosen arbitrarly.
Most of them, apart from CZK and UAH are a part of the [iShares Core MSCI World UCITS ETF](https://www.ishares.com/uk/individual/en/products/251882/ishares-msci-world-ucits-etf-acc-fund?switchLocale=y&siteEntryPassthrough=true). 
CZK and UAH has been added additionally, as both Czechia and Ukraine can be considered as EM.

```{r currencies_spec, warning=FALSE}
currencies <- c("BRL", "CLP", "MXN", "INR", "IDR","MYR","PHP", "KRW",
                "TWD", "THB", "PLN", "RUB", "CZK", "UAH", "ZAR", "TRY")
currencies <- tolower(currencies)
eur_currencies <- paste0(currencies,"eur")
usd_currencies <- paste0(currencies,"usd")

```

## Downloading data

Small function `read_data_stooq`{.r} enables to download all data from [stooq.pl](https://stooq.pl). Combining it with the `map`{.r} function from `purr`{.r} package allows to download as a list data for all currencies specified in `currencies`{.r} vector.

```{r currencies_download, warning=FALSE, message=FALSE}

data_all_eur <-read_data_stooq(eur_currencies)
data_all_usd <-read_data_stooq(usd_currencies)

```

# Preparing Data

Whole data contains data since 1995. For analyzing Coronavirus Crisis only data from 2020 is needed.

```{r data_spec_dates}

data_all_eur_2020 <- data_all_eur %>% 
  filter(Date > "2020-01-01")
data_all_usd_2020 <- data_all_usd %>% 
  filter(Date > "2020-01-01")

```

There are no NAs in the data, as they were extrapolated/intrapolated in the `read_stooq_data` function.

## Calculating Retuns

Let's calculate the returns - namely, log returns - and plot them. As, I am going to use the plot function for 2 different currencies, I will write a function for it.

```{r plot_function}
plot_log_returns_em_currencies <- function(input_data, currency_name){
  
  input_data %>%
  gather("currency", "price", -Date) %>% #Adapting for ggplot2 format
  ggplot(aes(x=Date,y=price)) +
  facet_wrap(~ currency, scales = "free") +
  geom_line() +
  theme_minimal() +
  ggtitle(paste("Log-Returns of EM currencies in 2020 over",currency_name))
  
}

```

Firstly, log-returns over EUR.

```{r generate_returns_eur}

data_all_eur_2020_lr <- generate_returns(data_all_eur_2020)
plot_log_returns_em_currencies(data_all_eur_2020_lr, "EUR")

```

And over USD.

```{r}

data_all_usd_2020_lr <- generate_returns(data_all_usd_2020)
plot_log_returns_em_currencies(data_all_usd_2020_lr, "USD")

```

# Return Analysis

> What is the mean return for EUR year to date?

```{r, echo=FALSE}
crisis_2020_eur_ret_mean <- data_all_eur_2020_lr %>% select(-Date)%>% summarise_all(lst(sum))
crisis_2020_eur_ret_mean %>% t() 
```

So, the most severly hit are BRL, MXN, RUB, ZAR and TRY.

> What is the mean return for USD yearto date?

```{r, echo=FALSE}
crisis_2020_usd_ret_mean <- data_all_usd_2020_lr %>% select(-Date)%>% summarise_all(lst(sum))
crisis_2020_usd_ret_mean %>% t() 
```

Similarly, BRL, MXN, RUB, ZAR and TRY are the worst hit currencies.

> Are any currencies getting stronger in relation to EUR than to USD?

```{r}

eurusd_com <- crisis_2020_eur_ret_mean[1,]/crisis_2020_usd_ret_mean[1,]
eurusd_com %>% t()

```

Only PHP and TWD performed better when demonimented in EUR terms rather than USD.
In both cases the appreciation over EUR was stronger, which makes sense, as the EUR got weaker in relation to USD in 2020.
All other currencies, lost more in value in USD than in EUR.
So, USD is still the king. Again.